Article Title

### 2014-11-28 manuscript.tex
corrections based on co-authors' inputs

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix} is
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
    \end{equation}
where $n$ is the number of time bins in $T$. The sample covariance matrix yields an unbiased estimate of the covariance matrix ($\E{C_{\sf sample}}=\Sigma$) if the mean, $\bar x$, is independent of the data sample, $\{x(t): t\in T$\}. When the mean is estimated from the same sample, $C_{\sf sample}$ becomes biased toward zero.  However, in all the cases where unbiasedness is required, we will estimate the mean from a separate sample.

@@

As the size of recorded neuronal populations increases, without substantial increases in recording durations, the \emph{condition number} of the sample covariance matrix also increases, making the partial correlation estimates \emph{ill-conditioned}: small errors in covariance estimates translate into much greater errors in partial correlations after the inversion. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization} \cite{Ledoit:2004,Schafer:2005}.

@@

Note that when the ground truth had zero correlations (Column A), $C_{\sf factor}$ performed equally well to $C_{\sf diag}$ because it correctly inferred zero factors and only estimated the individual variances. 
Similarly, when the number of latent units was zero (Column C), $C_{\sf sparse+latent}$ performed nearly equally well to $C_{\sf sparse}$ because it correctly inferred zero latent units.
With increasing sample sizes, all estimators converged to the truth (zero loss).  
However, the discriminability between their performances only increased with sample size (data not shown).

@@

Indeed, the validation loss computed by 10-fold cross-validation (see Methods) accurately reproduces the relative values of the true loss and the rankings of the covariance estimators without access to the ground truth (Fig.~\ref{fig:1} Row 6).

@@

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in the supragranular layers in primary visual cortex of lightly sedated mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:3} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:3} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:3} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:3} D). To estimate the noise correlation matrix, we presented only two distinct directions in some experiments or five directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:3} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:6} D).

@@

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (Fig.~\ref{fig:6} C). This suggests that correlations between neurons build up from multiple chains of smaller interactions. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting greater uniformity of average partial correlations than the smaller populations.

@@

\subsection*{Combining pairwise interactions and coactivations}
In the effort to account for the joint activity patterns that are poorly explained by pairwise interactions, investigators have augmented models of pairwise interactions with additional mechanisms such as latent variables, higher-order correlations, or global network fluctuations \cite{Ganmor:2011, Tkacik:2013, Pfau:2013, Koster:2014, Ecker:2014}.

@@

\subsection*{Addressing ill-posedness}
Inferring the conditional dependencies between variables in a probabilistic model is an ill-posed problem: small variations in the data produce large errors in the inferred network of dependencies. The problem becomes worse as the number of  recorded neurons increases until such models lose their statistical validity \cite{Roudi:2009}.  As techniques have improved to allow recording from larger neuronal populations, experimental neuroscientists have addressed this problem by extending the recording durations to keep sampling noise in check and verified that existing models are not overfitted \cite{Tkacik:2013}. However, ambitious projects, such as the BRAIN initiative  \cite{Alivisatos:2013}, aim to record from significantly larger populations. Simply increasing recording duration will not be sufficient, and the problem must be addressed by using regularized estimators. Regularization biases the solution toward a small subspace in order to counteract the effect of  sampling noise in the empirical data. However, biasing the solution to an inappropriate subspace does not allow significant estimation improvement and hinders interpretation.

@@

Distances between cells in physical space and in sensory feature space had a stronger effect on the partial correlations estimated by the $C_{\sf sparse+latent}$ estimator than on sample correlations (Fig.~\ref{fig:7} A--C).
These differences support the idea that correlations are built up from partial correlations in chains of intermediate cells positioned closer and tuned more similarly to one another, with potentially closer correspondence to anatomical connectivity.  These differences may also be at least partially explained by a trivial effect of regularization: the $L_1$ penalty applied by the estimator (Eq.~\ref{eq:ma}) suppresses small partial correlations to greater extent than large partial correlations, enhancing the apparent effect of distance and tuning.  Distances between cells in physical space and in sensory feature space also had distinct effects on the positive and negative connectivities inferred by the estimator (Fig.~\ref{fig:7} D--F); we could not think of a numerical artifact that could produce this effect. The distinct positive and negative connectivity patterns  may reflect geometric and graphical features of local excitatory and inhibitory networks. Indeed, the relationships between patterns of positive and negative connectivities inferred by the estimator mirrored the properties of excitatory and inhibitory synaptic connectivities with respect to distance, cortical layers, and feature tuning \cite{Song:2005, Oswald:2008, Adesnik:2010, Perin:2011, Fino:2011, Hofer:2011, Isaacson:2011, Levy:2012}. For example, while excitatory neurons form synapses within highly specific local cliques \cite{Perin:2011}, inhibitory interneurons form synapses with nearly all excitatory cells within local microcircuits \cite{Fino:2011, Hofer:2011, Packer:2011}.  To further investigate the link between synaptic connectivity and inferred functional connectivity, in future experiments, we will use molecular markers for various cell types with follow-up multiple whole-cell \emph{in vitro} recordings \cite{Hofer:2011, Ko:2013} to directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological functions. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections).

@@

Estimator $C_{\sf sparse+latent}$ (Eq.~\ref{eq:c-sl}) estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units.
Then the covariance matrix of the observed population is
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The rank of the $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ matches the number of the latent units in the joint distribution. Rather than finding $S_{12}$ and $S_{22}$ separately, $L$ can be estimated as a low-rank positive semidefinite matrix. To simultaneously optimize the sparse component $S$ and the low-rank component $L$, we adapted the loss function with $L_1$ penalty on $S$ combined with a penalty on the trace of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}\label{eq:ma}
(S,L) = \argmin\limits_{\hat S,\hat L} \loss{\hat S-\hat L, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta\Tr(\hat L)
\end{equation}
The trace of a symmetric semidefinite matrix equals the sum of the absolute values of its eigenvalues, \emph{i.e.} its \emph{nuclear norm}; penalty on $\Tr(L)$ favors solutions with few non-zero eigenvalues or, equivalently, low-rank solutions while keeping the convexity of the overall optimization problem \cite{Fazel:2002,Recht:2010}. This allows using convex optimization algorithm such as ADMM to be applied with great computational efficiency \cite{Ma:2013}.

>>>>

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is a $p\times 1$ vector of firing rates in time bin $t$, and an unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix} is
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
    \end{equation}
where $n$ is the number of time bins in $T$. 
The sample covariance matrix yields an unbiased estimate of the true covariance matrix ($\E{C_{\sf sample}}=\Sigma$) if the mean activity, $\bar x$, is estimated independently, in a separate data sample. 
In all cases when the unbiasedness of the sample covariance matrix matters in this paper, the means are estimated independently.

@@

Increasing the number of recorded neurons results in a higher \emph{condition number} of the covariance matrix \cite{Ledoit:2004} making the partial correlation estimates \emph{ill-conditioned}: small errors in the covariance estimates translate into much greater errors in the estimates of partial correlations. With massively multineuronal recordings, partial correlations cannot be estimated without \emph{regularization} \cite{Ledoit:2004,Schafer:2005}.

@@

Note that when the ground truth had zero correlations (Column A), $C_{\sf factor}$ performed equally well to $C_{\sf diag}$ because it correctly inferred zero factors and only estimated the individual variances. 
Similarly, when the number of latent units was zero (Column C), $C_{\sf sparse+latent}$ performed nearly equally well to $C_{\sf sparse}$ because it correctly inferred zero latent units.
With increasing sample sizes, all estimators converged to the ground truth (zero loss) but the estimators with correct structure outperformed the others even for large samples.

@@

Indeed, the validation loss computed by 10-fold cross-validation (see Methods) accurately reproduced the relative values of the true loss and the rankings of the covariance estimators without access to the ground truth (Fig.~\ref{fig:1} Row 6).

@@

\subsection*{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of densely sampled populations of neurons in layers 2/3 and upper layer 4 in primary visual cortex of sedated mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (Fig.~\ref{fig:3} A--B) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (Fig.~\ref{fig:3} C and D).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (Fig.~\ref{fig:3} C). Only cells that produced detectable calcium activity were included in the analysis (see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was displayed for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (Fig.~\ref{fig:3} D). To estimate the noise correlation matrix, we presented only two distinct directions in some experiments or five directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned on the stimulus onset for the estimation of the correlation matrix.   The sample correlation coefficients were largely positive and low (Fig.~\ref{fig:3} E and F). The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051 with the mean across sites of 0.018 (Fig.~\ref{fig:6} D).

@@

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 27 sites were about 5 times lower than the average sample correlations (Fig.~\ref{fig:6} C). This suggests that correlations between neurons build up from multiple chains of smaller interactions. Furthermore, the average partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.45 whereas that of the average partial correlations was 0.29, with larger populations exhibiting greater uniformity of average partial correlations than the smaller populations ($p=0.002$ Brown-Forsythe test).

@@

\subsection*{Combining pairwise interactions and coactivations}
In the effort to account for the joint activity patterns that are poorly explained by pairwise interactions, investigators have augmented models of pairwise interactions with additional mechanisms such as latent variables, higher-order correlations, or global network fluctuations \cite{Ganmor:2011, Tkacik:2013, Pfau:2013, Koster:2013, Ecker:2014}.

@@

\subsection*{Addressing ill-posedness}
Inferring the conditional dependencies between variables in a probabilistic model is an ill-posed problem: small variations in the data can produce very large errors in the inferred network of dependencies. The problem becomes worse as the number of  recorded neurons increases until such models lose their statistical validity \cite{Roudi:2009}.  As techniques have improved to allow recording from larger neuronal populations, experimental neuroscientists have addressed this problem by extending the recording durations to keep sampling noise in check and verified that existing models are not overfitted \cite{Tkacik:2013}. However, ambitious projects, such as the BRAIN initiative  \cite{Alivisatos:2013}, aim to record from significantly larger populations. Simply increasing recording duration will not be practical or sufficient, and the problem must be addressed by using regularized estimators. Regularization biases the solution toward a small subspace in order to counteract the effect of  sampling noise in the empirical data. However, biasing the solution to an inappropriate subspace does not allow significant estimation improvement and hinders interpretation.

@@

Distances between cells in physical space and in sensory feature space had a stronger effect on the partial correlations estimated by the $C_{\sf sparse+latent}$ estimator than on sample correlations (Fig.~\ref{fig:7} A--C).
These differences support the idea that correlations are built up from partial correlations in chains of intermediate cells positioned closer and tuned more similarly to one another, with potentially closer correspondence to anatomical connectivity.  These differences may also be at least partially explained by a trivial effect of regularization: the $L_1$ penalty applied by the estimator (Eq.~\ref{eq:ma}) suppresses small partial correlations to greater extent than large partial correlations, enhancing the apparent effect of distance and tuning.  
Still, the distinct positive and negative connectivity patterns (Fig.\ \ref{fig:7} D--F)  may reflect geometric and graphical features of local excitatory and inhibitory networks. 
Indeed, the relationships between patterns of positive and negative connectivities inferred by the estimator resembled the properties of excitatory and inhibitory synaptic connectivities with respect to distance, cortical layers, and feature tuning \cite{Song:2005, Oswald:2008, Adesnik:2010, Perin:2011, Fino:2011, Hofer:2011, Isaacson:2011, Levy:2012}. For example, while excitatory neurons form synapses within highly specific local cliques \cite{Perin:2011}, inhibitory interneurons form synapses with nearly all excitatory cells within local microcircuits \cite{Fino:2011, Hofer:2011, Packer:2011}.  To further investigate the link between synaptic connectivity and inferred functional connectivity, in future experiments, we will use molecular markers for various cell types with follow-up multiple whole-cell \emph{in vitro} recordings \cite{Hofer:2011, Ko:2013} to directly compare the inferred functional connectivity graphs to the underlying anatomical circuitry. Finally, the latent units inferred by the estimator can be analyzed for their physiological functions. For example, these latent units may be modulated under different brain states (e.g. slow-wave sleep, attention) and stimulus conditions (e.g. certain types of stimuli may engage feedback connections) \cite{Reimer:2014,Fu:2014}.

@@

Estimator $C_{\sf sparse+latent}$ (Eq.~\ref{eq:c-sl}) estimates a larger sparse precision matrix $S^\ast$ of the joint distribution of the $p$ observed neurons and $d$ latent units.
\begin{equation}
S^\ast=
\begin{pmatrix}
S & S_{12} \\
S_{12}^\T & S_{22}
\end{pmatrix},
\end{equation}
where the $p\times p$ partition $S$ corresponds to the visible units.
Then the covariance matrix of the observed population is
\begin{equation}
C_{\sf sparse+latent} = \left(S-S_{12}S_{22}^{-1}S_{12}^\T\right)^{-1}
\end{equation}
The rank of the $p\times p$  matrix $L=S_{12}S_{22}^{-1}S_{12}^\T$ matches the number of the latent units in the joint distribution. Rather than finding $S_{12}$ and $S_{22}$ separately, $L$ can be estimated as a low-rank positive semidefinite matrix. To simultaneously optimize the sparse component $S$ and the low-rank component $L$, we adapted the loss function with $L_1$ penalty on $S$ combined with a penalty on the trace of $L$ \cite{Chandrasekaran:2010,Ma:2013}:
\begin{equation}\label{eq:ma}
	(S,L) = \argmin\limits_{\hat S,\hat L} \loss{(\hat S-\hat L)^{-1}, C_{\sf sample}} + \alpha\|\hat S\|_1 + \beta\Tr(\hat L)
\end{equation}
The trace of a symmetric semidefinite matrix equals the sum of the absolute values of its eigenvalues, \emph{i.e.} its \emph{nuclear norm}; penalty on $\Tr(L)$ favors solutions with few non-zero eigenvalues or, equivalently, low-rank solutions while keeping the convexity of the overall optimization problem \cite{Fazel:2002,Recht:2010}. This allows using convex optimization algorithm such as ADMM to be applied with great computational efficiency \cite{Ma:2013}.
